---
title: "Portfolio optimisation using Genetic Algortithm"
author: "Sudeep Chakravarty"
date: "2026-02-12"
output: pdf_document
latex_engine: xelatex
---
# 1. Introduction

Portfolio optimisation is a fundamental problem in finance, where the objective
is to determine an optimal allocation of capital across a set of assets. Traditional
approaches such as the Markowitz mean-variance framework rely on convex optimisation
techniques. However, evolutionary algorithms such as Genetic Algortithms (GAs) provide
a flexible alternative capable of handling complex objective functions and constraints.

In this assignment, a Genetic Algortithm is used to optimise portfolio weights in order
to balance expected return and risk.The performance of the evolved portfolio is
evaluated on unseen future data and compared against equal-weighted and randomly 
generated portfolios. Additionally, different levels of risk prefrence are explored
using a lambda parameter, and the GA is extended to perform asset selection from a
larger pool.

# 2. Construction of Portfolio using GA

Portfolio Construction using Genetic Algorithm:

A set of 10 assets was selected and historical daily adjusted closing prices were 
obtained. Daily returns were calculated using log-return calculations. The optimisation
objective was defined using a risk-adjusted return functions:
  
Where lambda represents the investor's risk aversion parameter.

A real-valued Genetic Algorithm was implemented using the GA package. The algorithm evolves candidate portfolio weights under the constraint that all weights are non-negative and sum
to one.The main parameter used were:
  
  - Population size:50
  - Maximum generations:100
  - Real-valued encoding
  - Fitness maximisation

The convergence plot shows that the fitness value stabilises after several generations, indicating that the GA successfully converged to a stable optimal solution.

In this implemention, the objectives was to maximize expected return while 
penalising risk throught the lambda parameter. A higher lambda increases the
weight given to portfolio volatility.

```{r,echo=FALSE, message=FALSE, warning=FALSE}
library(quantmod)
library(PerformanceAnalytics)
library(GA)

# Select 10 assets
tickers <- c("AAPL", "MSFT", "GOOG", "AMZN", "META","TSLA", "NVDA", "JPM", "NFLX", "AMD")
# Download Data(Training)
getSymbols(tickers, from = "2019-01-01", to = "2023-12-31")
# Extract adjusted Close prices
prices <- do.call(merge, lapply(tickers, function(x) Cl(get(x))))
colnames(prices) <- tickers

# Calculate daily returns
returns <- na.omit(Return.calculate(prices))
head(returns)

# Fitness Function define
fitness_function <- function(w,returns){
w <- w / sum(w)
port_return <- mean(returns %*% w)
port_risk <- sd(returns %*% w)
lambda <- 0.5
fitness <- port_return - lambda * port_risk
return(fitness)
}

# GA Run
num_assets <- ncol(returns)
ga_result <-ga(type = "real-valued",
fitness = function(w)
fitness_function(w, returns),
                lower = rep(0, num_assets),
                upper = rep(1, num_assets),
                popSize = 50,
                maxiter = 100,
                run = 50)
                
# Optimal Weights calculate
optimal_weights <- as.numeric(ga_result@solution)
optimal_weights <- optimal_weights / sum(optimal_weights)
optimal_weights

#  Portfolio Performance calcul5ate
opt_port_return <- mean(returns %*% optimal_weights)
opt_port_risk <- sd(returns %*% optimal_weights)
opt_port_return
opt_port_risk

cat("Optimal Weights:\n")
round(optimal_weights, 4)

cat("In-sample Expected Return:", round(opt_port_return, 5), "\n")
cat("In-sample Risk:", round(opt_port_risk, 5), "\n")

# Equal weights setup (used later)
n_assets <- ncol(returns)
equal_weights <- rep(1/n_assets, n_assets)

# Plot GA goverenance
plot(ga_result)
```

# 3. Evaluation on unseen Future Data

Evaluation on Unseen data:

To assess generalisation performance, the dataset was divided into training and
testing periods. The GA was trained on historical data(training period), and the
optimised weights were than applied to unseen future data(test period).

The evolved portfolio maintained reasonable performance on the unseen dataset.
Althought some performance variation was observed, the portfolio continued to
demonstrate a balanced return-risk profile. This suggests that the GA did not
signficantly overfit the training data.

Splitting the dataset into training and testing periods helps reduce the risk of
overfitting. A model that performs well only on training data may not generalise 
to unseen data. The comparable performance observed on the test period suggests
that the GA solution captures meaningful structure in the data rather than noise.

```{r, echo=FALSE}
# Equal weights setup (used later)
n_assets <- ncol(returns)
equal_weights <- rep(1/n_assets, n_assets)

# Split data
train_returns <- returns["2019/2022",]
test_returns <- returns["2023",]

# Weights normalized
optimal_weights <- as.numeric(ga_result@solution)
optimal_weights <- optimal_weights / sum(optimal_weights)
optimal_weights

# Performance on test data
test_port_return <- mean(test_returns %*% optimal_weights)
test_port_risk <- sd(test_returns %*% optimal_weights)
test_port_return
test_port_risk
cat("Out-of-sample Expected Return:", round(test_port_return, 5), "\n")
cat("Out-of-sample Risk:", round(test_port_risk, 5), "\n")
ga_test_returns <- test_returns %*% optimal_weights
equal_test_returns <- test_returns %*% equal_weights

ga_cum <- cumprod(1 + ga_test_returns)
equal_cum <- cumprod(1 + equal_test_returns)

plot(index(test_returns),
     ga_cum,
     type = "l",
     col = "red",
     lwd = 2,
     xlab = "Date",
     ylab = "Cumulative Growth",
     main = "Out-of-Sample Performance (2023)")

lines(index(test_returns),
      equal_cum,
      col = "blue",
      lwd = 2)

legend("topleft",
       legend = c("GA Optimized", "Equal Weight"),
       col = c("red", "blue"),
       lwd = 2)
```

# 4. Comparison with Evenly Weighted and Random Portfolio

Comparison with Equal-Weighted and Random Portfolios:

To evaluate the effectiveness of the GA-based optimisation,the evolved portfolio 
was compared with:
  
  1. An equally weighted portfolio.
  2. A randomly generated portfolio.

Results shows that the GA-optimised portfolio achieves a superior risk-return
tradeoff. While the equal-weighted portfolio provides a stable baseline, and the
random portfolio consistently delivers improved return per unit as risk.

This indicates that evolutionary optimisations is capable of identifying more 
efficient alloctions compared to native allocations strategies.

The GA-optimized portfolio achives a higher expected return with competitive
risk compared to equal-weighted and random portfolios, demonstrating the
effectiveness of evolutionary optimisation.

The equally weighted portfolio serves as a simple and widely used benchmark, as
it does not rely on optimisation assumptions. The random portfolio provide a
baseline for naive allocation. The GA based approach outperforming these
benchmark indicates that the optimisation process adds value beyond naive
diversification.

```{r,echo=FALSE}
# Equal weights setup (used later)
n_assets <- ncol(returns)
equal_weights <- rep(1/n_assets, n_assets)
# Evenly Weighted Portfolio
n_assets <- ncol(returns)
equal_weights <- rep(1/n_assets, n_assets)
equal_port_return <- mean(returns %*% equal_weights)
equal_port_risk <- sd(returns %*% equal_weights)

# Random Portfolio

set.seed(123)
random_weights <- runif(n_assets)
random_weights <- random_weights / sum(random_weights)
random_port_return <- mean(returns %*% random_weights)
random_port_risk <- sd(returns %*% random_weights)

comparison <- data.frame(
Portfolio = c("GA Optimized", "Equal Weighted", "Random"),
Return = c(opt_port_return, equal_port_return, random_port_return),
Risk = c(opt_port_risk, equal_port_risk, random_port_risk)
)
comparison
colors <- c("red", "blue", "darkgreen")

plot(comparison$Risk,
     comparison$Return,
     pch = 19,
     col = colors,
     cex = 1.5,
     xlab = "Portfolio Risk (Standard Deviation)",
     ylab = "Expected Return",
     main = "Risk vs Return Comparison")

text(comparison$Risk,
     comparison$Return,
     labels = comparison$Portfolio,
     pos = 4)

legend("bottomright",
       legend = comparison$Portfolio,
       col = colors,
       pch = 19)
```

# 5. Portfolios with Differntly Balanced Risk and Return

Risk-Return Trade With Different Lambda Values:

  To simulate different investor prefrence, the optimisation was repeated for
  multiple values of lambda.
  
  - Lower lambda places more emphasis on maximising return.
  - Higher lambda places greater emphasis on minimising risk.
  
  the result demonstrate a clear risk-return tradeoff. As lambda increases, portfolio
  risk decreases but expected return also declines. This behaviour aligns with
  financial theory and resembles the concept of an efficient frontier.
  
  These findings confirm that the GA framework can flexibly adapt to varying 
  levels of risk tolerance.
    
  As lambda increases, the optimisation objective increasingly penalises portfolio
  volatility, resulting in lower risk allocations and adjusted expected returns.
  This illustrates the classical risk–return trade-off in evolutionary 
  optimisation.
  
  From an economic perspective, investors with low risk aversion (low lambda)
  may prefer higher-return but more volatile portfolios. Conversely, conservative
  investors may choose higher lambda values to prioritise capital preservations.
  This demonstrates the flexibility of the GA framework in adapting to different
  investor profiles.
  
```{r,echo=FALSE}
lambda_values <- c(0,5, 1,2)
results_lambda <- data.frame(
Lambda = numeric(),
Return = numeric(),
Risk = numeric()
)
for (l in lambda_values) {
fitness_function <- function(w) {
w <- w / sum(w)
port_return <- mean(returns %*% w)
port_risk <- sd(returns %*% w)
return(port_return - 1 * port_risk)
}
ga_result <- ga(type = "real-valued", fitness = fitness_function,
                                      lower = rep(0, ncol(returns)),
                                      upper = rep(1, ncol(returns)),
                                      popSize = 50,
                                      maxiter = 100,
                                      run = 50)
optimal_weights <- as.numeric(ga_result@solution)
optimal_weights <- optimal_weights / sum(optimal_weights)

port_return <- mean(returns %*% optimal_weights)
port_risk <- sd(returns %*% optimal_weights)

results_lambda <- rbind(results_lambda, data.frame(Lambda = l,
Return = port_return,
Risk = port_risk))
}
results_lambda
plot(results_lambda$Risk,
     results_lambda$Return,
     pch = 19,
     col = "purple",
     xlab = "Risk",
     ylab = "Return",
     main = "Risk-Return Tradeoff for Different Lambda")

text(results_lambda$Risk,
     results_lambda$Return,
     labels = results_lambda$Lambda,
     pos = 4)

risk_values <- c(opt_port_risk, equal_port_risk, random_port_risk)
return_values <- c(opt_port_return, equal_port_return, random_port_return)
plot(risk_values, return_values,
xlab="Risk(Std Dev)")
ylab="Return (Mean)"
main="Risk vs Return Comparison"
pch=19
col=c("blue", "red", "darkgreen")
text(risk_values, return_values, labels=c("GA", "Equal", "Random"),
pos=4)
```
# 6. Asset Selection using Genetic Algorithm

 Assets Selections using GA:
  
  The GA was further extended to perform asset selection from a larger pool of
  assets. Instead of only optimising weights, the algorithm also determined which
  assets should be included in the portfolio.
  
  This increases the solution space significantly, as the GA mush now search both
  the assets subset and th weights allocations simultaneosly.
  
  The selected subset demonstrated competitive performance relative to the full
  portfolio. This suggests that the GA is capable of identifying a smaller set of
  high-performing assets while maintaining favorable risk-return characteristics.
  
  Assets selection using evolutionary methods provides an advantages over manual
  selection by exploring a much larger combination space effeciently.
  
The genetic algorithm selected AAPL, MSFT, GOOG & JPM as the optimal subset of 
assets, indicating that these stocks contributed most effectively to the 
risk-return objective.

The selected assets include large technology and financial firms, which may offer
strong return potential combined with diversification benefits. The GA appears
to favour assets that contribute positively to the overalll risk-adjusted 
objective. This suggests that the algorithm effectively balances individual assets 
performance with portfolio-level risk characteristics.

```{r,echo=FALSE}
# Fitness function for asset selection
fitness_selection <- function(binary_vector){
if (sum(binary_vector) == 0) {
return(-999)
}
# Selected assets for subsets
selected_returns <- returns[, as.logical(binary_vector)]
# Equal weighted amoung selected assets
weights <- rep(1/ ncol(selected_returns), ncol(selected_returns))
# Portfolio return & risk
port_return <- mean(selected_returns %*% weights)
port_risk <- sd(selected_returns %*% weights)
# Objective: maximize return - risk
fitness_value <- port_return - port_risk
return(fitness_value)
}
# Binary Generic Algorithm
ga_selection <- ga(type = "binary", fitness = fitness_selection, 
nBits = ncol(returns),
popSize = 50,
maxiter = 100,
run = 50)
# solution
best_selection <- ga_selection@solution
# Convert to logical selection
selected_assets <- tickers[as.logical(best_selection)]
selected_assets
```
# 7. Conclusion

Conclusion:

  This study demonstrates that Genetic Algortithms provide a powerful and flexible
  framework for portfolio optimisation. The GA successfully identifies optimal
  optimal weight allocations and showed stable convergence behaviour.
  
  While evaluated on unseen future data, the optimised portfolio maintained 
  competitive performance. Comparisons with equal-weighted and random portfolios
  confirmed the superiority of the evolutionary approach.
  
  Furthermore, varying the risk preference parameter illustracted a clear
  risk-return tradeoff, and extending the GA to assets selection further 
  enhanced the flexibility of the framework.
  
  Overall, evolutionary computations offers a robust alternative to traditional 
  portfolio optimisation techniques.
  
  Despite promising results, the GA approach has limitations. The optimisations 
  outcome may depend on parameter choices such as population size and number of
  generations. Additionally, financial markets are dynamic, and past data may not
  fully represent future conditions. Future work could explore alternative 
  objective functions or multi-objective optimisations techniques.
  